---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2022"
author: "Tatiana Sokolova"
subtitle: Assignment 4 - Due date 02/17/22
output:
  word_document: default
  pdf_document: default
geometry: margin=2.54cm
editor_options:
  chunk_output_type: inline
---
  
## Directions
  
  You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A04_Sp21.Rmd"). Submit this pdf using Sakai. 

R packages needed for this assignment: "xlsx" or "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r setup, include=FALSE} 
#setup for knit
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

```{r}
library(xlsx)
library(readxl)
library(lubridate)
library(ggplot2)
library(forecast)
library(tseries)
library(Kendall)
library(dplyr) 
```

## Questions 

Consider the same data you used for A3 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2021 Monthly Energy Review. For this assignment you will work only with the column "Total Renewable Energy Production". 


```{r}
#Importing data set - using xlsx package
energy_data <- read.xlsx(file="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",header=FALSE, startRow=13, sheetIndex=1)
#extracting column names from row 11
read_col_names <- read.xlsx(file="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",header=FALSE,startRow=11,endRow=11,sheetIndex=1)
colnames(energy_data) <- read_col_names
head(energy_data)

#creating df structure for column of interest and 
df <- energy_data[,c('Month','Total Renewable Energy Production')]
head(df)

#removing January 1973 to compare with differenced df
df_584<-df[-c(1),]
head(df_584)
```


## Stochastic Trend and Stationarity Tests

### Q1

Difference the "Total Renewable Energy Production" series using function diff(). Function diff() is from package base and take three main arguments:
* *x* vector containing values to be differenced;
* *lag* integer indicating with lag to use;
* *differences* integer indicating how many times series should be differenced.

Try differencing at lag 1 only once, i.e., make `lag=1` and `differences=1`. Plot the differenced series. Do the series still seem to have trend?

```{r}
energy_data_diff <- diff(df[,"Total Renewable Energy Production"],lag=1, differences=1)

ggplot(df_584, aes(x=df_584[,1], y=df_584[,"Total Renewable Energy Production"])) +
  geom_line(aes(x=df_584[,1], y = energy_data_diff, color = "Differenced")) +
  geom_line(aes(x=df_584[,1], y=df_584[,"Total Renewable Energy Production"], color = "Original")) +
  labs(color="") +
  scale_color_manual(values = c("Differenced" = "green", "Original" = "black"),
                                labels=c("Differenced", "Original")) +
  theme(legend.position = "bottom") +
  ggtitle("Total Renewable Energy Production")+
  ylab(paste0("Energy Production (Trillion Btu)")) +
  xlab(paste0("Month"))
      
```
The series appears to be detrended.

### Q2

Now let's compare the differenced series with the detrended series you calculated on A3. In other words, for the "Total Renewable Energy Production" compare the differenced series from Q1 with the series you detrended in A3 using linear regression. (Hint: Just copy and paste part of your code for A3)

Copy and paste part of your code for A3 where you compute regression for Total Energy Production and the detrended Total Energy Production

```{r}
#Create vector t
nobs=nrow(df)
t <- c(1:nobs)

#Fit a linear trend to TS of Total Renewable Energy Production
linear_trend_model_renew=lm(df[,2]~t)  
summary(linear_trend_model_renew)
beta0r=as.numeric(linear_trend_model_renew$coefficients[1])  #first coefficient is the intercept term or beta0
beta1r=as.numeric(linear_trend_model_renew$coefficients[2])  #second coefficient is the slope or beta1

#remove the trend from TS of Total Renewable Energy Production
detrend_renew_data <- df[,2]-(beta0r+beta1r*t)

```

### Q3
Create a data frame with 4 columns: month, original series, detrended by Regression Series and differenced series. Make sure you properly name all columns. Also note that the differenced series will have only 584 rows because you lose the first observation when differencing. Therefore, you need to remove the first observations for the original series and the detrended by regression series to build the new data frame.

```{r}
#removing Jan 1973 from detrended data
detrend_renew_data <- detrend_renew_data[-1]
head(detrend_renew_data)

new_df <- data.frame(Month = df_584$Month,
                             Original = df_584[,2],
                             Detrended = detrend_renew_data, 
                             Differenced = energy_data_diff)
head(new_df)
```

### Q4
Using ggplot() create a line plot that shows the three series together. Make sure you add a legend to the plot.

```{r}
#Use ggplot
ggplot(new_df, aes(x=new_df[,1], y=new_df[,2])) +
  geom_line(aes(x=new_df[,1], y= new_df[,2], color = "Original")) +
  geom_line(aes(x=new_df[,1], y = new_df[,3], color = "Detrended")) +
  geom_line(aes(x=new_df[,1], y= new_df[,4], color = "Differenced")) +
  labs(color="") +
  scale_color_manual(values = c("Original" = "black","Detrended" = "blue","Differenced" = "green"),
                                labels=c("Original", "Detrended", "Differenced")) +
  theme(legend.position = "bottom") +
  ggtitle("Total Renewable Energy Production")+
  ylab(paste0("Energy Production (Trillion Btu)")) +
  xlab(paste0("Month"))
      
```


### Q5
Plot the ACF for the three series and compare the plots. Add the argument `ylim=c(-0.5,1)` to the Acf() function to make sure all three y axis have the same limits. Which method do you think was more efficient in eliminating the trend? The linear regression or differencing?

```{r}
#Comparing ACFs
par(mar=c(3,3,3,0));par(mfrow=c(1,3))
Acf(new_df$Original,lag.max=60,main="Original",ylim=c(-0.5,1))
Acf(new_df$Detrended,lag.max=60,main="Detrended",ylim=c(-0.5,1))
Acf(new_df$Differenced,lag.max=60,main="Differenced",ylim=c(-0.5,1))
```

The differencing was more efficient in eliminating the trend.

### Q6
Compute the Seasonal Mann-Kendall and ADF Test for the original "Total Renewable Energy Production" series. Ask R to print the results. Interpret the results for both tests. What's the conclusion from the Seasonal Mann Kendall test? What's the conclusion for the ADF test? Do they match what you observed in Q2? Recall that having a unit root means the series has a stochastic trend. And when a series has stochastic trend we need to use a different procedure to remove the trend.

```{r}
#converting to time series
ts_renew_energy <- ts(df[,2],frequency=12)

#Seasonal Mann-Kendall
SMKtest <- SeasonalMannKendall(ts_renew_energy)
print("Results for Seasonal Mann Kendall")
print(summary(SMKtest)) 

```
The p value is sharing how significant the tau value is and it definitely statistically different from 0 so the null hypothesis that there is no trend present in the data should be rejected. The larger score indicates that it is a positive trend. This matches what I observed in Q2.

```{r}
#Null hypothesis is that data has a unit root
print("Results for ADF test")
print(adf.test(ts_renew_energy,alternative = "stationary")) #stationary over a unit root but could be non stationary 

```
From the ADF test, since the p-value is greater than 0.05, we can conclude that we fail to reject the null hypothesis that the time series is non-stationary and can therefore conclude the time series has some time-dependent structure and does not have a constant variance over time. This matches what I observed in Q2.

### Q7

Aggregate the original "Total Renewable Energy Production" series by year. You can use the same procedure we used in class. Store series in a matrix where rows represent months and columns represent years. And then take the columns mean using function colMeans(). Recall the goal is the remove the seasonal variation from the series to check for trend. 

```{r}
energy_data_matrix <- matrix(df[,2],byrow=FALSE,nrow=12)
energy_data_yearly <- colMeans(energy_data_matrix)

#for Spearman test
my_year <- c(year(first(df[,1])):year(last(df[,1])))
head(my_year)
```

### Q8

Apply the Mann-Kendall, Spearman correlation rank test, and ADF. Are the results from the test in agreement with the test results for the non-aggregated series, i.e., results for Q6?

```{r}
#Mann-Kendall
print("Results of Mann Kendall on average yearly series")
print(summary(MannKendall(energy_data_yearly)))

```
The p-value is still statistically different from 0 so the rejection of the null hypothesis (there is no trend present in the data) still holds. Tau is slightly bigger than that of the non-aggregated series, implying that there is an even stronger positive correlation within this aggregated series.

```{r}
#Spearman Correlation Test
print("Results from Spearman Correlation")
sp_rho=cor.test(energy_data_yearly,my_year,method="spearman") 
print(sp_rho)
```
Question 6 did not require the Spearman test because it cannot handle seasonality. The p-value for the test on the yearly data indicates that there is very strong evidence for rejecting the null hypothesis that there is no monotonic association in the series. The rho indicates that there is a strong positive correlation within this aggregated series.

```{r}
#Augmented Dickey-Fuller Test
print("Results for ADF test on yearly data")
print(adf.test(energy_data_yearly, alternative = "stationary"))
```
The p-value is still greater than 0.05 and therefore we can continue to conclude that the time series is stationary. The Dickey-Fuller value is more negative than that of the non-aggregated series, implying that there is an even stronger rejection of the null hypothesis.
